Income and Food Price Transmission

### 2016-08-11 analysis.Rnw
JAE sub. vers.

% !Rnw root = master.Rnw

@@

\subsubsection{Domestic and international food prices}
Data on domestic prices are sourced from the International Labor Organization (ILO).\footnote{See \citet{ilo16}} ILO maintains a comprehensive database of domestic food price indexes (FPIs) and consumer price indexes (CPIs) that is used to monitor the cost of living and real wages. There are not many studies which draw on this database, however, despite a broad country coverage and long monthly price series. A major issue with this data is that historical series are not recalculated whenever a new base year is introduced. The same applies when other changes are made such as the inclusion or exclusion of certain items such as beverages or tobacco from the index. This means that many of the series have discrete level shifts associated with these changes. In some cases one series replace another after a period of overlap. Finally, many of the series contain missing values. To ``tidy up'' the raw data I carried out the following steps First, I restricted the sample to the period 2005:1-2014:12. Secondly, in cases where multiple complete series were available from the same country, a single one was selected. Thirdly, series with missing values were discarded. Fourthly, and most critical for the analysis below, level shifts associated with base year changes etc. were identified and removed.\footnote{Concretely I added the change in the break point from the previous value to the break point and all points succeeding it. The value of the adjusted series at the break point is thus the same as the one preceding it.} Finally, all the series were normalized with respect to a common base period (Jan 2007). In total \Sexpr{n.full} countries have complete price series for the entire ten year period considered whereas, for example, there are \Sexpr{n.sub} countries with complete price series covering the 2005-11 subperiod. There is therefore a trade-off between the number of complete price series and the length of the series.

@@

International food prices are represented by the FAO food price index (FAO FPI) and in some cases its ``real'' version, namely the nominal FAO FPI deflated by the Manufactures Unit Value (MUV) index published by the World Bank.\footnote{see \citet{fao16} and \citet{worldbank16a}.}

@@

There has been two recent international price spikes; one in 2007-8 and the other in 2010-11. The former spike especially is often referred to as ``the food crisis'' in the literature \citep[e.g.][]{Headey2008, Headey2013, timmer10, Ivanic2012, Martin2011}. Between January 2007 and June 2008 the FAO FPI rose by 91 points to index 226, corresponding to a 68 percent relative price increase. From the first peak in mid-2008 and to the trough in February 2009 the FAO FPI fell by 82 points before rising back up to index 240 in February 2011.

@@

\subsubsection{Price transmission determinants}
Table \ref{tab:cs_data} in the appendix lists the variables that, based on the discussion in section \ref{sec:model}, are considered potentially important determinants of price transmission. A country's PPP adjusted per capita GDP is considered a proxy for its income level. Other covariates considered are a country's percentage dietary energy supply derived from cereals, roots and tubers, its cereal import- and export shares, a dummy indicating whether the country is a landlocked developing country (LLDC), its trade openness defined as the sum of its imports and exports relative to its GDP and a dummy indicating whether there is a grain state trading enterprise (STE) in the country. In addition there are two indexes included representing the regulatory performance of the country in question and the quality of its infrastructure. These two indexes are abbreviated DTF (Distance to Frontier) and LPI (Logistics Performance Index), respectively.\footnote{The covariates are similar to those of \citet{greb12}.}

\subsection{Determinants of domestic food price changes during the food crisis\label{sec:analysis1}}
Table \ref{tab:07-08-fc-regressions} considers the cross sectional variation in domestic food price changes during the 2007-8 food crises. Each of the columns refer to a regression of FPI inflation or real FPI growth on the explanatory variables discussed above as well as the percentage change in a country's domestic currency/US dollar exchange rate during the food crisis. All variables except for GDP and the dummies have been standardized in order to facilitate comparisons of effect sizes.

@@

A concern is that some of the variables are highly correlated which leads to large standard errors and insignificant effects. Therefore, the subset of covariates included in columns (2) and (4) are those which minimizes the Akaike information criterion (AIC) as determined by stepwise regression.\footnote{This was carried out using the \emph{MASS} package in \textsf{R} \citep{venables02}} Stepwise regressions have been criticized for leading to models where the remaining terms appear to be more important than they actually are. However, as can be seen, no substantial changes result from this exercise.

What effect did the government interventions have on domestic food inflation during the food crisis? Table \ref{tab:int-regressions} in the appendix reports results from a set of regressions similar to those in table \ref{tab:07-08-fc-regressions} except that a set of dummies are included indicating whether or not the government intervened in various ways during the food crisis.\footnote{These dummies are based on table 1 in \citet{Demeke2009}. ``Export'' = 1 if the country Restricted or banned export, ``Price'' = 1 if the country administered price control or restricted private trade, ``Stock'' = 1 if the country released stocks (public or imported) at subsidized prices'', ``Tariff'' = 1 if the country reduced tariffs or customs fees on imports and ``Tax'' = 1 if the country suspended or reduced VAT or other taxes, all as of 1. December 2008.\label{int_note}} Some of the coefficients to these intervention dummies are substantial but they are all statistically insignificant. Table \ref{tab:1011-regressions} in the appendix contains results related to the 2010-11 international price spike. Again, estimated effects are similar to those in  table \ref{tab:07-08-fc-regressions} but most are statistically insignificant. In fact,  the cereal consumption share is the only robustly significant predictor of domestic food inflation in 2010-11.

@@

Prior to estimation I tested each of the series for unit roots and cointegration with international prices.\footnote{I tested for stationarity with the KPSS test \citep{kwiatkowski92} and for unit roots with the augmented Dickey-Fuller test \citep{dickey79}. I used the Phillips and Ouliaris ``Pz'' test \citep{phillips90} to test for cointegration. These tests are implemented in \textsf{R} via the packages \emph{tseries} \citep{trapletti15} and \emph{urca} \citep{pfaff08}, respectively.} Results indicated that the domestic FPI and real FPI series, are nonstationary in levels but stationary in first differences. Domestic price levels were also found not to be cointegrated with world prices. Consequently, each model was estimated in growth rates rather than levels in order to avoid spurious regressions.

@@

The LRM is a common measure of price transmission. \cite{IMF2011}, for example, contains an analysis of food price price transmission based on a dataset of monthly FPI inflation from 31 advanced economies and 47 emerging and developing economies over the period 2000 - 2011. Their median estimated LRM pass-through of a 1 percent increase in international food inflation is 0.18 percent in developed countries and 0.34 percent in emerging and developing countries. Similarly, \citet{Ianchovichina2014} report an average estimated LRM for the eighteen Middle East and North Africa (MENA) countries of 0.25 and the majority of the countries are reported to have  food price pass-through LRMs in the order of 0.2-0.4. Both studies use 12 lags in their country specific ADL models, i.e. they assume $p=q=12$.\footnote{\cite{IMF2011} do not include exchange rates in their regressions.} My own estimates based on a sample of \Sexpr{n} countries covering the period 2005:1 to 2014:12 are very similar. The average and median FPI LRM from a model with 12 lags of all variables are \Sexpr{round(fivenum.fpi["Mean", "V12"], 2)} and \Sexpr{round(fivenum.fpi["Median", "V12"], 2)}, respectively. That being said, it is not obvious that so many lags should be included in the models. When lag length selection is based on the Akaike Information Criterion (AIC), for example, the optimal number lags in most of the models turn out to be $p=q=1$.\footnote{For each country, I calculated the AIC for each of the $12^2$ combinations of $(p,q)\, \in (1,..,12)$. Then I selected the $(p,q)$ combination with the lowest AIC.} But when a common lag length of 1 is used for all countries, average and median FPI transmission LRMs are much lower, namely \Sexpr{round(fivenum.fpi["Mean", "V1"], 2)} and \Sexpr{round(fivenum.fpi["Median", "V1"], 2)}, respectively, whereas the average of the standard errors of the LRM estimates very similar.\footnote{I also tested for residual autocorrelation with the Durbin-Watson test. There was no evidence of residual autocorrelation in either case. Standard errors of the LRMs are based on the so called Bewley transformation of the ADL model instrumented with lags of the dependent variable.} As a compromise between the long lag lengths used in the literature which might cause problems with overfitting and the short optimal lag lengths implied by the AIC criterion which leads to LRM estimates that are lower than commonly found in the literature, the country specific models are based on common lag lengths of $p=q=6$. This choice leads to average long run FPI and RFPI transmission rates of \Sexpr{round(fivenum.fpi["Mean", "V6"], 2)} and \Sexpr{round(fivenum.rfpi["Mean", "V6"], 2)}, respectively. One interpretation of the very low RFPI LRM average is that shocks to a country's FPI is passed on almost fully to its non-FPI in the long run, a phenomenon known as propagation in the literature \citep[e.g.][]{pedersen10}.

The long run transmission estimates, $\hat{\theta}$, are regressed, on the same explanatory variables as in table \ref{tab:07-08-fc-regressions}. However, since the dependent variable is an estimate rather than an observed value and the model now represents an estimated dependent variable (EDV) regression \citep[see][]{hanushek74, lewis05}. A common approach in this situation is to use Weighted Least Squares (WLS) estimation with weights $w_{i}$ that are inversely proportional to the standard error of $\theta_{i}$ \citep{saxonhouse76}. This way more precise estimates are given a higher weight in the regression. The main reason though for using WLS rather than OLS is that the latter may suffer from heteroskedasticity which leads to inefficient coefficient estimates inconsistent standard errors. The inconsistent OLS standard errors can, however, easily be corrected with a robust standard error estimator. \citet{lewis05} uses formal reasoning backed up by a Monte Carlo analysis to argue that OLS with robust standard errors is likely to produce more reliable estimates than WLS based on inverse standard error weights, and that the latter is likely to produce downward biased standard errors.

Table \ref{tab:fpi-pt-regressions} presents estimates of FPI transmission determinants based on OLS [columns (1)-(2)] as well as WLS [columns (3)-(4)], both with robust standard errors. Again, all variables except per capita GDP and the dummies have been standardized and, as in table \ref{tab:fpi-pt-regressions}, column (1) and (3) represent regressions that are based on the full set of covariates whereas columns (2) and (4) represent regressions with the lowest AIC.

@@

Table \ref{tab:rfpi-pt-regressions} in the appendix contain regression results based on estimated long run RFPI transmission rates from country specific time series models with 6 lags. The signs of the coefficients are similar to those in table \ref{tab:fpi-pt-regressions} but most effects are statistically insignificant. Similarly, table \ref{tab:fpi-pt1-regressions} and \ref{tab:rfpi-pt1-regressions} in the appendix contain results based on FPI and real FPI transmission rates calculated from time series models with a single lag. These also point to only a few significant predictors of price transmission. A plausible reason is that the LRM estimates based on models with a single lag are relatively more noisy and therefore more difficult to explain.

@@

It is well known that dynamic panel models such as equation (\ref{eq:panel_model}) suffers from (Nickell) bias due to correlation between the lagged dependent variable and the error term. However, as demonstrated by \citet{nickell81}, the bias of the within estimator decreases at a rate $1/T$, i.e. endogeneity is mainly a problem in short panel datasets with only a few time periods $T$. In the present case where $T>100$, the bias should be negligible.

@@

<<echo=false>>=
load("tp_ctr_mean.Rda")
load("tp_ctr_mean1.Rda")
@

\subsection{Estimated income-price transmission turning points}
Table \ref{tab:tp} summarizes the estimated income turning points derived from the significant coefficients in table \ref{tab:07-08-fc-regressions}-\ref{tab:panel-regressions}. The ratios in the last column refer to the number of countries with a per capita GDP less than the estimated turning point shown in the second-last column, relative to the total number of countries in the sample with data on income. The average share of countries that are to the left of the turning point, across all specifications, is \Sexpr{round(nc.pc.mean[1],0)} percent. There are, however, important differences between the different estimators. Panel estimation, in particular, leads to a significantly higher proportion of countries to the left of the turning point than cross section estimation (\Sexpr{round(nc.pc.mean[2],0)} vs. \Sexpr{round(nc.pc.mean[3],0)} percent for the non-panel estimators) on average. In constant 2011 international dollars, the average income level in 2014 among the group of low-, lower middle- and upper middle income countries was \$1,517, \$5,752 and \$13,719, respectively \citep{worldbank16b}.\footnote{According to the World Bank classification scheme. MICs are defined as having a 2011 per capita gross national income of US\$1,026 to \$12,475, where the conversion from local currency to US dollars is based on the Atlas method.} In comparison, the income level leading to the highest predicted price transmission, averaged across all specifications, is \$\Sexpr{format(round(tp.mean[1],0), big.mark= ",", scientific= FALSE)}. Countries from the lower end of the high-income spectrum experience the highest short run food price transmission whereas lower middle income countries experience the highest long run food price transmission according to these estimates.

>>>>

% !Rnw root = Master.Rnw

@@

\subsubsection{Domestic and international food prices}
Data on domestic prices are sourced from the International Labor Organization (ILO). ILO maintains a comprehensive database of domestic food price indexes (FPIs) and consumer price indexes (CPIs) that is used to monitor the cost of living and real wages. Not many studies which draw on this database, however, despite a broad country coverage and long monthly price series. A major issue with this data is that historical series are not recalculated whenever a new base year is introduced. The same applies when other changes are made such as the inclusion or exclusion of certain items such as beverages or tobacco from the index. This means that many of the series have discrete level shifts associated with these changes. In some cases one series replace another after a period of overlap. Finally, many of the series contain missing values. To ``tidy up'' the raw data I carried out the following steps: First, I restricted the sample to the period 2005:1-2014:12. Secondly, in cases where multiple complete series were available from the same country, a single one was selected. Thirdly, series with missing values were discarded. Fourthly, and most critical for the analysis below, level shifts associated with base year changes etc. were identified and removed.\footnote{Concretely I added the change in the break point from the previous value to the break point and all points succeeding it. The value of the adjusted series at the break point is thus the same as the one preceding it.} Finally, all the series were normalized with respect to a common base period (Jan 2007). In total \Sexpr{n.full} countries have complete price series for the entire ten year period considered whereas, for example, there are \Sexpr{n.sub} countries with complete price series covering the 2005-11 subperiod. There is therefore a trade-off between the number of complete price series and the length of the series.

@@

International food prices are represented by the FAO food price index (FAO FPI) and in some cases its ``real'' version, namely the nominal FAO FPI deflated by the Manufactures Unit Value (MUV) index published by the World Bank.

@@

There has been two recent international price spikes; one in 2007-8 and the other in 2010-11. The former spike especially is often referred to as ``the food crisis'' in the literature \citep[e.g.][]{Headey2008, Headey2013, Timmer2010, Ivanic2012, Martin2011}. Between January 2007 and June 2008 the FAO FPI rose by 91 points to index 226, corresponding to a 68 percent relative price increase. From the first peak in mid-2008 and to the trough in February 2009 the FAO FPI fell by 82 points before rising back up to index 240 in February 2011.

@@

\subsubsection{Price transmission determinants}
Table \ref{tab:cs_data} in the appendix lists the variables that, based on the discussion in section \ref{sec:model}, are considered potentially important determinants of price transmission. A country's PPP adjusted per capita GDP is considered a proxy for its income level. Other covariates considered are a country's percentage dietary energy supply derived from cereals, roots and tubers, its cereal import- and export shares, a dummy indicating whether the country is a landlocked developing country (LLDC), its trade openness defined as the sum of its imports and exports relative to its GDP and a dummy indicating whether there is a grain state trading enterprise (STE) in the country. In addition there are two indexes included representing the regulatory performance of the country in question and the quality of its infrastructure. These two indexes are abbreviated DTF (Distance to Frontier) and LPI (Logistics Performance Index), respectively.\footnote{The covariates are similar to those of \citet{Greb2016}.}

\subsection{Determinants of domestic food price changes during the food crisis\label{sec:analysis1}}
Table \ref{tab:07-08-fc-regressions} considers the cross sectional variation in domestic food price changes during the 2007-8 food crises. Each of the columns refer to a regression of FPI inflation or real FPI growth on all or a subset of the explanatory variables discussed above as well as the percentage change in a country's domestic currency/US dollar exchange rate during the food crisis. All variables except for GDP and the dummies have been standardized to facilitate comparisons of effect sizes.

@@

A concern is that some of the variables are highly correlated which leads to large standard errors and insignificant effects. This problem is amplified by the loss of observations due to missing data on some of these variables for some countries. Therefore, columns (2) and (4) consider only the variable of primary interest, namely income. As can be seen, the increased number of observations and the reduced number of explanatory variables decreases $R^2$ substantially whereas the significance of the income effects remains or increases.

What effect did the government interventions have on domestic food inflation during the food crisis? Table \ref{tab:int-regressions} in the appendix reports results from a set of regressions similar to those in table \ref{tab:07-08-fc-regressions} except that a set of dummies are included indicating whether or not the government intervened in various ways during the food crisis.\footnote{These dummies are based on table 1 in \citet{Demeke2009}. ``Export'' = 1 if the country Restricted or banned export, ``Price'' = 1 if the country administered price control or restricted private trade, ``Stock'' = 1 if the country released stocks (public or imported) at subsidized prices'', ``Tariff'' = 1 if the country reduced tariffs or customs fees on imports and ``Tax'' = 1 if the country suspended or reduced VAT or other taxes, all as of 1. December 2008.\label{int_note}} Some of the coefficients to these intervention dummies are substantial but they are mainly statistically insignificant. Table \ref{tab:1011-regressions} in the appendix contains results related to the 2010-11 international price spike. Again, estimated effects are similar to those in  table \ref{tab:07-08-fc-regressions}. The main difference is that cereal consumption is a significant predictor of food inflation in this case.

@@

Prior to estimation I tested each of the series for unit roots and cointegration with international prices.\footnote{I tested for stationarity with the KPSS test \citep{Kwiatkowski1992} and for unit roots with the augmented Dickey-Fuller test \citep{Said1984}. I used the Phillips and Ouliaris ``Pz'' test \citep{Phillips1990} to test for cointegration. These tests are implemented in \textsf{R} via the packages \emph{tseries} \citep{Trapletti2007} and \emph{urca} \citep{Pfaff2005}, respectively.} Results indicated that the domestic FPI and real FPI series, are nonstationary in levels but stationary in first differences. Domestic price levels were also found not to be cointegrated with world prices. Consequently, each model was estimated in growth rates rather than levels in order to avoid spurious regressions.

@@

The LRM is a common measure of price transmission. \cite{IMF2011}, for example, contains an analysis of food price price transmission based on a dataset of monthly FPI inflation from 31 advanced economies and 47 emerging and developing economies over the period 2000 - 2011. Their median estimated LRM pass-through of a 1 percent increase in international food inflation is 0.18 percent in developed countries and 0.34 percent in emerging and developing countries. Similarly, \citet{Ianchovichina2014} report an average estimated LRM for the eighteen Middle East and North Africa (MENA) countries of 0.25 and the majority of the countries are reported to have  food price pass-through LRMs in the order of 0.2-0.4. Both studies use 12 lags in their country specific ADL models, i.e. they assume $p=q=12$.\footnote{\cite{IMF2011} do not include exchange rates in their regressions.} My own estimates based on a sample of \Sexpr{n} countries covering the period 2005:1 to 2014:12 are very similar. The average and median FPI LRM from a model with 12 lags of all variables are \Sexpr{round(fivenum.fpi["Mean", "V12"], 2)} and \Sexpr{round(fivenum.fpi["Median", "V12"], 2)}, respectively. That being said, it is not obvious that so many lags should be included in the models. When lag length selection is based on the Akaike Information Criterion (AIC), for example, the optimal number lags in most of the models turn out to be $p=q=1$.\footnote{For each country, I calculated the AIC for each of the $12^2$ combinations of $(p,q)\, \in (1,..,12)$. Then I selected the $(p,q)$ combination with the lowest AIC.} But when a common lag length of 1 is used for all countries, average and median FPI transmission LRMs are much lower, namely \Sexpr{round(fivenum.fpi["Mean", "V1"], 2)} and \Sexpr{round(fivenum.fpi["Median", "V1"], 2)}, respectively, whereas the average of the standard errors of the LRM estimates are very similar.\footnote{I also tested for residual autocorrelation with the Durbin-Watson test. There was no evidence of residual autocorrelation in either case. Standard errors of the LRMs are based on the so called Bewley transformation of the ADL model instrumented with lags of the dependent variable.} As a compromise between the long lag lengths used in the literature which might cause problems with overfitting and the short optimal lag lengths implied by the AIC criterion which leads to LRM estimates that are lower than commonly found in the literature, the country specific models are based on common lag lengths of $p=q=6$. This choice leads to average long run FPI and RFPI transmission rates of \Sexpr{round(fivenum.fpi["Mean", "V6"], 2)} and \Sexpr{round(fivenum.rfpi["Mean", "V6"], 2)}, respectively. One interpretation of the very low RFPI LRM average is that shocks to a country's FPI is passed on almost fully to its non-FPI in the long run, a phenomenon known as propagation in the literature \citep[e.g.][]{Pedersen2010}.

The long run transmission estimates, $\hat{\theta}$, are regressed, on the same explanatory variables as in table \ref{tab:07-08-fc-regressions}. However, since the dependent variable is an estimate rather than an observed value and the model now represents an estimated dependent variable (EDV) regression \citep[see][]{Hanushek1974, Lewis2005}. A common approach in this situation is to use Weighted Least Squares (WLS) estimation with weights $w_{i}$ that are inversely proportional to the standard error of $\theta_{i}$ \citep{Saxonhouse1976}. This way more precise estimates are given a higher weight in the regression. The main reason though for using WLS rather than OLS is that the latter may suffer from heteroskedasticity which leads to inefficient coefficient estimates inconsistent standard errors. The inconsistent OLS standard errors can, however, easily be corrected with a robust standard error estimator. \citet{Lewis2005} uses formal reasoning backed up by a Monte Carlo analysis to argue that OLS with robust standard errors is likely to produce more reliable estimates than WLS based on inverse standard error weights, and that the latter is likely to produce downward biased standard errors.

Table \ref{tab:fpi-pt-regressions} presents estimates of FPI transmission determinants based on OLS [columns (1)-(2)] as well as WLS [columns (3)-(4)], both with robust standard errors. Again, all variables except per capita GDP and the dummies have been standardized and, as in table \ref{tab:fpi-pt-regressions}, column (1) and (3) represent regressions that are based on the full set of covariates whereas columns (2) and (4) represent regressions based solely on income.

@@

Table \ref{tab:rfpi-pt-regressions} in the appendix contain regression results based on estimated long run RFPI transmission rates from country specific time series models with 6 lags. The signs of the coefficients are similar to those in table \ref{tab:fpi-pt-regressions} but most effects are statistically insignificant.

@@

It is well known that dynamic panel models such as equation (\ref{eq:panel_model}) suffers from (Nickell) bias due to correlation between the lagged dependent variable and the error term. However, as demonstrated by \citet{Nickell1981}, the bias of the within estimator decreases at a rate $1/T$, i.e. endogeneity is mainly a problem in short panel datasets with only a few time periods $T$. In the present case where $T>100$, the bias should be negligible.

@@

<<echo=false>>=
str = "C:/Users/nqw235/Google Drive/Research/pov_price_transmission/"
load(paste0(str,"tp_ctr_mean.Rda"))
load(paste0(str,"tp_ctr_mean1.Rda"))
@

\subsection{Estimated income-price transmission turning points}
Table \ref{tab:tp} summarizes the estimated income turning points derived from the significant coefficients in table \ref{tab:07-08-fc-regressions}-\ref{tab:panel-regressions}. The ratios in the last column refer to the number of countries with a per capita GDP less than the estimated turning point shown in the second-last column, relative to the total number of countries in the sample with data on income. The average share of countries that are to the left of the turning point, across all specifications, is \Sexpr{round(nc.pc.mean[1],0)} percent. There are, however, important differences between the different estimators. Panel estimation, in particular, leads to a significantly higher proportion of countries to the left of the turning point than cross section estimation (\Sexpr{round(nc.pc.mean[2],0)} vs. \Sexpr{round(nc.pc.mean[3],0)} percent for the non-panel estimators) on average. According to the World Bank, the average income level in 2014 among the group of low-, lower middle- and upper middle income countries was \$1,517, \$5,752 and \$13,719, respectively, when measured in constant 2011 international dollars.\footnote{According to the World Bank classification scheme. MICs are defined as having a 2011 per capita gross national income of US\$1,026 to \$12,475, where the conversion from local currency to US dollars is based on the Atlas method.} In comparison, the income level leading to the highest predicted price transmission, averaged across all specifications, is \$\Sexpr{format(round(tp.mean[1],0), big.mark= ",", scientific= FALSE)}. Countries from the lower end of the high-income spectrum experience the highest short run food price transmission whereas lower middle income countries experience the highest long run food price transmission according to these estimates.
